A Constructive Ordered Predicate Logic \\ for Category Theory

### 2022-10-14 paper/paper.tex
fix refs

First, we using the elimination for the unit set, we can see that all
constructions are (pro-)functorial:
\begin{construction}
  For any small category $\cat C$, we can construct natural elements
  \begin{enumerate}
  \item Identity: $\pendallXdotY {\alpha} {\punitinXfromYtoZ {\cat C}{\alpha}{\alpha'}}$
  \item Composition: $\hombinary {\alpha_1} {\punitinXfromYtoZ {}{\alpha_1}{\alpha_2}} {\alpha_2} {\punitinXfromYtoZ {}{\alpha_1}{\alpha_2}} {\alpha_3} {\punitinXfromYtoZ {}{\alpha_1}{\alpha_3}}$
  \item Functoriality: for any $F : \varr {\cat C}{\cat D}$, $\homunary {\alpha_1} {\punitinXfromYtoZ {}{\alpha_1}{\alpha_2}} {\alpha_2} {\punitinXfromYtoZ {} {F(\alpha_1)}{F(\alpha_2)}}$.
   \item Profunctoriality: for any $R : \harr {\cat C}{\cat D}$ if
     $\cat D$ is small then \\ $\homtrinary {\alpha_1}
     {\punitinXfromYtoZ{} {\alpha_1}{\alpha_2}} {\alpha_2}
     {R{\alpha_2}{\beta_2}} {\beta_2} {\punitinXfromYtoZ{} {\beta_2}
       {\beta_1}} {\beta_1} {R\alpha_1\beta_1}$
  \end{enumerate}
\end{construction}
Reflexivity and transitivity generalize the corresponding properties
of equality, with the lack of symmetry being a key feature of the
generalization.
%
In addition, we can prove that the (pro)-functoriality axioms commute
with the composition proof by the $\eta$ principle for the unit.
%
(Pro-)Functoriality generalizes the statement that all functions and
relations respect equality.
%
Naturality is more complex to state, and it is a statement about the
\emph{proofs} so it has no analogue in ordinary higher-order
logic. The following version is stated for any \emph{profunctor}, with
the usual case of naturality arising when $R \alpha \beta =
\punitinXfromYtoZ {\cat C}{F \alpha}{G \beta}$.
\begin{lemma}[Naturality]
  For any $t : \pendallXdotY {\alpha:\cat C}{\harrapp R \alpha
    \alpha}$, by composing with profunctoriality, we can construct
  terms $\alpha_1, f : \punitinXfromYtoZ {\cat C} {\alpha_1}
  {\alpha_2}, \alpha_2 \vdash \textrm{lcomp}(f,\pendappXtoY t
  {\alpha_2})$ and $\textrm{rcomp}(\pendappXtoY t {\alpha_1}, f) :
  \harrapp R {\alpha_1}{\alpha_2}$ that are both equal to
  $\punitelimtoYwithkontZ f {t}$.
\end{lemma}

@@

    \inferrule*[right=ElementSubstExt]
    {\Gamma\pipe\Psi \vdash \phi :: \Phi \and
      d^+\Psi = d^-\Psi'
     \Gamma\pipe\Psi' \vdash t : R[d^+\phi;b/\beta]\and
     \Gamma\pipe d^+\Psi'\vdash b : \cat D}
    {\Gamma\pipe\Psi \vdash \phi,t/x,b/\beta :: (\Phi,x:R,\beta:\cat D)}
  \end{mathpar}
  \caption{Substitution}
  \label{fig:substitions}
\end{figure}

@@

\begin{definition}
  \label{lem:vertical-sq-cat}
  We define the vertical composition of transformation substitutions
  $\phi[\psi]$ inductively on $\phi$.
  \begin{align*}
    (a/\alpha)[b/\beta] &= a[b/\beta]/\alpha\\
    (\phi_1,t/x,a/\alpha)[\psi_1\jnctx\psi_2] &= \phi_1[\psi_1], t[\psi_2],a[d^+\psi_2]
  \end{align*}
  This covers all cases by \cref{inversion}.

@@

\begin{theorem}[Adjunctions]
  $G$ is right adjoint to $F$ if and only if there is a natural isomorphism
  $\punitinXfromYtoZ {\cat D} {(F \alpha)} \beta \cong \punitinXfromYtoZ {\cat D} \alpha {G \beta}$
\end{theorem}
\begin{proof}
  First, by \cref{lem:co-unit}, the forward direction of the
  isomorphism is equivalent to providing a term
  \[ \eta : \pendallXdotY \alpha \punitinXfromYtoZ \alpha {G(F(\alpha)) } \]

  Second, the backward direction of the isomorphism is equivalent to
  providing a term
  \[ \epsilon : \pendallXdotY \beta \punitinXfromYtoZ {F(G(\beta))} \beta \]

  And the triangle laws state exactly that the induced functions are
  isomorphisms.
\end{proof}

\subsection{Kan Extensions/Lifts}

As a further example of the synthetic approach to universal
properties, we consider right and left \emph{Kan extensions}, which
are generalizations of limits and colimits, respectively.

The right Kan extension of $F$ along $J$ is the ``best'' (maximal) way to
fill in the following square:
\[% https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAGEQBfU9TXfIRQBGclVqMWbACLdeIDNjwEiZYePrNWiDgHJu4mFADm8IqABmAJwgBbJGRA4ISAEzVNUnQCk5lm-aIjs5IohJabABiBlxAA
\begin{tikzcd}
\cat C \arrow[d, "J"] \arrow[r, "F"] & \cat D \\
\cat C'                              &  
\end{tikzcd}\]
I.e. a functor $\textrm{Ran}_J F : \cat C' \to \cat D$ with a
universal natural transformation $\textrm{Ran}_J F \circ J \Rightarrow
F$. A left kan extension $\textrm{Lan}_J F$ is the same but with a
natural transformation $F \Rightarrow \textrm{Lan}_J F \circ J $.

We can express the concept in SCT as representing a certain profunctor
constructed using the profunctor hom:
\begin{definition}[Kan Extensions]
  Given functors $F : \varr {\cat C}{\cat D}$ and $J : \varr {\cat
    C}{\cat C'}$, a right Kan extension is a functor $\textrm{Ran}_J F
  : \varr {\cat C'}{\cat D} $that right-represents the profunctor
  \[ \hlambda {\beta:\cat D}{\alpha': \cat C} \homrallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat C'}{\alpha'}{J\alpha})} {\punitinXfromYtoZ {\cat D}{\beta}{F \alpha}} \]

  A left Kan extensions is a functor $\textrm{Lan}_J F : \varr {\cat
    C'}{\cat D}$ is a functor satisfying the dual property: it
  \emph{left}-represents the profunctor
  \[\hlambda {\alpha':\cat C}{\beta:\cat D} \homlallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat C'}{J\alpha}{\alpha'})} {\punitinXfromYtoZ {\cat D}{F \alpha}{\beta}}
  \]
\end{definition}

The limit of a functor $F : \varr{\cat C}{\cat D}$ is a special case
of a Kan extension where $J : \varr{\cat C}{1}$ is the unique functor
to the terminal category. Similarly, the colimit of $F$ is the left
Kan extension with respect to the same $J$.
%
Specializing to this case, the representability for limits means:
\[ \punitinXfromYtoZ {\cat D}{\beta}{\textrm{lim} F} \cong \homrallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat 1}{\alpha'}{()})}{\punitinXfromYtoZ {\cat D}{\beta}{F \alpha}} \]
And if we view a morphism in $\cat 1$ as being trivial, then this
essentially states that mapping $\beta$ to $\textrm{lim} F$ is the
same thing as constructing a cone: for all $\alpha :\cat C$, a
morphism from $\beta$ to $F \alpha$, naturally in $\alpha$.

The usual formulation of Kan extensions is in terms of a universal
morphism.
%
We can reproduce this formulation using our established theory of
equivalences for homs of profunctors.
\begin{lemma}
  One half of the representability isomorphism for a right Kan
  extension $\textrm{Ran}_J F$ is determined by a natural
  transformation
  \[ \pendallXdotY \alpha \punitinXfromYtoZ {\cat D}{(\textrm{Ran}_JF)(J\alpha)}{F\alpha} \]

  Dually, one half of the representability isomorphism for a left Kan
  extension $\textrm{Lan}_J F$ is determined by a natural
  transformation
  \[ \pendallXdotY \alpha \punitinXfromYtoZ {\cat D}{F\alpha}{(\textrm{Lan}_JF)(J\alpha)} \]
\end{lemma}
\begin{proof}
  By \cref{lem:unit-repr}, one half of representability is determined by a natural element
  \[ \pendallXdotY {\alpha'} \homrallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat C'}{\alpha'}{J\alpha})} {\punitinXfromYtoZ {\cat D}{(\textrm{Ran}_J F)\alpha'}{F \alpha}} \]
  which, by \cref{lem:fubini} is isomorphic to
  \[
  \pendallXdotY {\alpha} \homlallXYtoZ {\alpha'}{(\punitinXfromYtoZ {\cat C'}{\alpha'}{J\alpha})}{\punitinXfromYtoZ {\cat D}{(\textrm{Ran}_J F)\alpha'}{F \alpha}}
  \]
  which, by \cref{lem:yo-and-co-yo}, is isomorphic to 
  \[ \pendallXdotY \alpha \punitinXfromYtoZ {\cat D}{(\textrm{Ran}_JF)(J\alpha)}{F\alpha} \]
  by taking $P = \hlambda {\alpha'}{\alpha} {\punitinXfromYtoZ {\cat D}{(\textrm{Ran}_J F)\alpha'}{F \alpha}}$.

  The proof for left Kan extensions follows by \cref{lem:duality}.
\end{proof}

>>>>

First, we using the elimination for the unit set, we can see that all
constructions are (pro-)functorial:
\begin{construction}
  \label{construction:synthetic-composition}
  For any small category $\cat C$, we can construct natural elements
  \begin{enumerate}
  \item Identity: $\pendallXdotY {\alpha} {\punitinXfromYtoZ {\cat C}{\alpha}{\alpha'}}$
  \item Composition: $\hombinary {\alpha_1} {\punitinXfromYtoZ {}{\alpha_1}{\alpha_2}} {\alpha_2} {\punitinXfromYtoZ {}{\alpha_1}{\alpha_2}} {\alpha_3} {\punitinXfromYtoZ {}{\alpha_1}{\alpha_3}}$
  \item Functoriality: for any $F : \varr {\cat C}{\cat D}$, $\homunary {\alpha_1} {\punitinXfromYtoZ {}{\alpha_1}{\alpha_2}} {\alpha_2} {\punitinXfromYtoZ {} {F(\alpha_1)}{F(\alpha_2)}}$.
   \item Profunctoriality: for any $R : \harr {\cat C}{\cat D}$ if
     $\cat D$ is small then \\ $\homtrinary {\alpha_1}
     {\punitinXfromYtoZ{} {\alpha_1}{\alpha_2}} {\alpha_2}
     {R{\alpha_2}{\beta_2}} {\beta_2} {\punitinXfromYtoZ{} {\beta_2}
       {\beta_1}} {\beta_1} {R\alpha_1\beta_1}$
  \end{enumerate}
\end{construction}
Reflexivity and transitivity generalize the corresponding properties
of equality, with the lack of symmetry being a key feature of the
generalization.
%
In addition, we can prove that the (pro)-functoriality axioms commute
with the composition proof by the $\eta$ principle for the unit.
%
(Pro-)Functoriality generalizes the statement that all functions and
relations respect equality.
%
Naturality is more complex to state, and it is a statement about the
\emph{proofs} so it has no analogue in ordinary higher-order
logic. The following version is stated for any \emph{profunctor}, with
the usual case of naturality arising when $R \alpha \beta =
\punitinXfromYtoZ {\cat C}{F \alpha}{G \beta}$.
\begin{lemma}[Naturality]
  For any $t : \pendallXdotY {\alpha:\cat C}{\harrapp R \alpha
    \alpha}$, by composing with profunctoriality, we can construct
  terms $\alpha_1, f : \punitinXfromYtoZ {\cat C} {\alpha_1}
  {\alpha_2}, \alpha_2 \vdash \textrm{lcomp}(f,\pendappXtoY t
  {\alpha_2})$ and $\textrm{rcomp}(\pendappXtoY t {\alpha_1}, f) :
  \harrapp R {\alpha_1}{\alpha_2}$ that are both equal to
  $\punitelimtoYwithkontZ f {t}$.
\end{lemma}

@@

    \inferrule*[right=ElementSubstExt]
    {\Gamma\pipe\Psi \vdash \phi :: \Phi \and
      d^+\Psi = d^-\Psi'
     \Gamma\pipe\Psi' \vdash t : R[d^+\phi;b/\beta]\and
     \Gamma\pipe d^+\Psi'\vdash b : \cat D}
    {\Gamma\pipe\Psi \vdash \phi,t/x,b/\beta :: (\Phi,x:R,\beta:\cat D)}
  \end{mathpar}
  \caption{Substitution}
  \label{fig:substitutions}
\end{figure}

@@

\begin{definition}
  \label{lem:vertical-sq-cat}
  We define the vertical composition of transformation substitutions
  $\phi[\psi]$ inductively on $\phi$.
  \begin{align*}
    (a/\alpha)[b/\beta] &= a[b/\beta]/\alpha\\
    (\phi_1,t/x,a/\alpha)[\psi_1\jnctx\psi_2] &= \phi_1[\psi_1], t[\psi_2],a[d^+\psi_2]
  \end{align*}
  This covers all cases by \cref{lem:inversion}.

@@

%% \begin{theorem}[Adjunctions]
%%   $G$ is right adjoint to $F$ if and only if there is a natural isomorphism
%%   $\punitinXfromYtoZ {\cat D} {(F \alpha)} \beta \cong \punitinXfromYtoZ {\cat D} \alpha {G \beta}$
%% \end{theorem}
%% \begin{proof}
%%   First, by \cref{lem:co-unit}, the forward direction of the
%%   isomorphism is equivalent to providing a term
%%   \[ \eta : \pendallXdotY \alpha \punitinXfromYtoZ \alpha {G(F(\alpha)) } \]

%%   Second, the backward direction of the isomorphism is equivalent to
%%   providing a term
%%   \[ \epsilon : \pendallXdotY \beta \punitinXfromYtoZ {F(G(\beta))} \beta \]

%%   And the triangle laws state exactly that the induced functions are
%%   isomorphisms.
%% \end{proof}

%% \subsection{Kan Extensions/Lifts}

%% As a further example of the synthetic approach to universal
%% properties, we consider right and left \emph{Kan extensions}, which
%% are generalizations of limits and colimits, respectively.

%% The right Kan extension of $F$ along $J$ is the ``best'' (maximal) way to
%% fill in the following square:
%% \[% https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAGEQBfU9TXfIRQBGclVqMWbACLdeIDNjwEiZYePrNWiDgHJu4mFADm8IqABmAJwgBbJGRA4ISAEzVNUnQCk5lm-aIjs5IohJabABiBlxAA
%% \begin{tikzcd}
%% \cat C \arrow[d, "J"] \arrow[r, "F"] & \cat D \\
%% \cat C'                              &  
%% \end{tikzcd}\]
%% I.e. a functor $\textrm{Ran}_J F : \cat C' \to \cat D$ with a
%% universal natural transformation $\textrm{Ran}_J F \circ J \Rightarrow
%% F$. A left kan extension $\textrm{Lan}_J F$ is the same but with a
%% natural transformation $F \Rightarrow \textrm{Lan}_J F \circ J $.

%% We can express the concept in SCT as representing a certain profunctor
%% constructed using the profunctor hom:
%% \begin{definition}[Kan Extensions]
%%   Given functors $F : \varr {\cat C}{\cat D}$ and $J : \varr {\cat
%%     C}{\cat C'}$, a right Kan extension is a functor $\textrm{Ran}_J F
%%   : \varr {\cat C'}{\cat D} $that right-represents the profunctor
%%   \[ \hlambda {\beta:\cat D}{\alpha': \cat C} \homrallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat C'}{\alpha'}{J\alpha})} {\punitinXfromYtoZ {\cat D}{\beta}{F \alpha}} \]

%%   A left Kan extensions is a functor $\textrm{Lan}_J F : \varr {\cat
%%     C'}{\cat D}$ is a functor satisfying the dual property: it
%%   \emph{left}-represents the profunctor
%%   \[\hlambda {\alpha':\cat C}{\beta:\cat D} \homlallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat C'}{J\alpha}{\alpha'})} {\punitinXfromYtoZ {\cat D}{F \alpha}{\beta}}
%%   \]
%% \end{definition}

%% The limit of a functor $F : \varr{\cat C}{\cat D}$ is a special case
%% of a Kan extension where $J : \varr{\cat C}{1}$ is the unique functor
%% to the terminal category. Similarly, the colimit of $F$ is the left
%% Kan extension with respect to the same $J$.
%% %
%% Specializing to this case, the representability for limits means:
%% \[ \punitinXfromYtoZ {\cat D}{\beta}{\textrm{lim} F} \cong \homrallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat 1}{\alpha'}{()})}{\punitinXfromYtoZ {\cat D}{\beta}{F \alpha}} \]
%% And if we view a morphism in $\cat 1$ as being trivial, then this
%% essentially states that mapping $\beta$ to $\textrm{lim} F$ is the
%% same thing as constructing a cone: for all $\alpha :\cat C$, a
%% morphism from $\beta$ to $F \alpha$, naturally in $\alpha$.

%% The usual formulation of Kan extensions is in terms of a universal
%% morphism.
%% %
%% We can reproduce this formulation using our established theory of
%% equivalences for homs of profunctors.
%% \begin{lemma}
%%   One half of the representability isomorphism for a right Kan
%%   extension $\textrm{Ran}_J F$ is determined by a natural
%%   transformation
%%   \[ \pendallXdotY \alpha \punitinXfromYtoZ {\cat D}{(\textrm{Ran}_JF)(J\alpha)}{F\alpha} \]

%%   Dually, one half of the representability isomorphism for a left Kan
%%   extension $\textrm{Lan}_J F$ is determined by a natural
%%   transformation
%%   \[ \pendallXdotY \alpha \punitinXfromYtoZ {\cat D}{F\alpha}{(\textrm{Lan}_JF)(J\alpha)} \]
%% \end{lemma}
%% \begin{proof}
%%   By \cref{lem:unit-repr}, one half of representability is determined by a natural element
%%   \[ \pendallXdotY {\alpha'} \homrallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat C'}{\alpha'}{J\alpha})} {\punitinXfromYtoZ {\cat D}{(\textrm{Ran}_J F)\alpha'}{F \alpha}} \]
%%   which, by \cref{lem:fubini} is isomorphic to
%%   \[
%%   \pendallXdotY {\alpha} \homlallXYtoZ {\alpha'}{(\punitinXfromYtoZ {\cat C'}{\alpha'}{J\alpha})}{\punitinXfromYtoZ {\cat D}{(\textrm{Ran}_J F)\alpha'}{F \alpha}}
%%   \]
%%   which, by \cref{lem:yo-and-co-yo}, is isomorphic to 
%%   \[ \pendallXdotY \alpha \punitinXfromYtoZ {\cat D}{(\textrm{Ran}_JF)(J\alpha)}{F\alpha} \]
%%   by taking $P = \hlambda {\alpha'}{\alpha} {\punitinXfromYtoZ {\cat D}{(\textrm{Ran}_J F)\alpha'}{F \alpha}}$.

%%   The proof for left Kan extensions follows by \cref{lem:duality}.
%% \end{proof}
